Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with \textit{ANTsR}

### 2014-05-03 paper.tex
ENH:  More edits.

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{K.~L.~Shrinidhi}
\author[label1]{Max Wintermark}
\author[label1]{Christopher R.~Durst}
\author[label1]{James C.~Gee}
\author[label3]{Murray C.~Grossman}
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, 
                 Department of Radiology, University of Pennsylvania,
                Philadelphia, PA}
\address[label3]{Department of Neurology, University of Pennsylvania,
                Philadelphia, PA}

@@

\begin{abstract} 
Automated brain tumor segmentation in MRI is complicated by the lack
of prior knowledge concerning tumor location, spatial extent, shape,
possible displacement of normal tissue, and intensity signature.
These factors limit the applicability of techniques with established
performance in related tasks such as normal brain tissue
segmentation.  To alleviate the limitations of traditional tissue
classification, we introduce a symmetric template-based framework for supervised
segmentation based on multiple modality intensity, geometry and asymmetry feature sets.  These features drive a supervised whole-brain and tumor segmentation
approach based on random forest-derived probabilities.  The
asymmetry-related features, in particular, 
demonstrate excellent discriminative properties within this framework. 
We also gain performance by alternately generating
probability maps from random forest models and feeding these maps 
into Markov random field regularized probabilistic segmentation.  This
strategy allows us to interface the supervised learning capabilities of the random forest
model with regularized probabilistic segmentation.  The newly
developed \textit{ANTsR} open source software package facilitates
the proposed methodology.  \textit{ANTsR} is a comprehensive interface between the
popular Advanced Normalization Tools (ANTs) and the \textit{R}
statistical project. \textit{ANTsR} integrates proven image analysis methods
for image registration, segmentation, bias correction, and
template building with \textit{R}'s statistical and machine
learning techniques and state-of-the-art visualization possibilities.  
The reported algorithmic framework was the
top-performing entry in the MICCAI 2013 Multimodal Brain Tumor
Segmentation challenge and, to our knowledge, is the only entry for
both challenge years (2012 \& 2013) which has been made publicly
available.
\end{abstract}

@@

\section{Introduction}
Given the complexity of tumor growth and appearance and the need
for precise volumetric measurements for tumor characterization, 
much research has been invested in computational approaches for 
automated segmentation of tumor regions in MR images.  Several 
approaches have been previously proposed in the literature as
detailed in recent reviews \citep{angelini2007,bauer2013}.  
Commensurable algorithmic evaluation, however, is
extremely problematic due to widely varying performance assessments
described in the corresponding publications, lack of publicly available 
evaluation data, and private algorithmic instantiations.  
In response to these issues, the Multimodal Brain Tumor Segmentation 
(BRATS) challenge was initiated in 2012 (and continued in 2013) under 
the auspices of the Medical Image Computing and Computer Assisted 
Intervention Society in association with their annual international 
conference \citep{menze2014}.

After observing the 2012 challenge to gain insight for our own research 
needs \citep{durst2014}, we found that no teams had availed their software in any 
form.  This prompted our participation in the challenge which took place
the following year in Nagoya, Japan.  Heavily influenced by the success 
of participants in the 2012 challenge (specifically, the work of 
\cite{bauer2012,geremia2012,zikic2012}), we adopted the increasingly popular 
random forest (RF) machine learning framework \citep{breiman2001} which 
permits the inclusion of many potentially discriminative image features.  
Supervised segmentation techniques generally consist of a training phase
for model construction using image-based feature data followed by prediction using the 
generated model.  For supervised brain tumor segmentation, 
a set of training data consisting of labeled brain images 
is used to construct a predictive model.  Although other 
classification techniques have been used to segment
brain tumors (e.g., support vector machines \citep{bauer2011}),
RF models have proven particularly successful.

Several machine learning concepts were integrated to create 
the RF framework first articulated in its entirety by Breiman
et al. \citep{breiman2001} for performing classification/regression.  
Although decision trees had been previously explored in the literature, 
it was the success of ``boosting''-style machine learning 
techniques, such as AdaBoost \citep{schapire1990,freund1997}, which influenced 
the aggregation of such decision trees into ``forests'' 
with randomized node optimization for improved
classification/regression performance \citep{ho1995,amit1997}.
The final element of bootstrap aggregating or ``bagging'' (i.e.
random sampling of the training data) was
introduced by Breiman \citep{breiman1996} to achieve improved
accuracy.%
\footnote{
One of the principal advantages of \textit{R} is the extensive community of
developers  who have contributed on the order of thousands of packages 
extending \textit{R} 's capabilities beyond its core functionality.
Most relevant here
is the {\tt randomForest} package developed from Breiman's original
Fortran code by Liaw and Wiener \citep{liaw2002}.
}

Early adoption \cite{viola2005} and success in the
computer vision community
has led to a recent surge within the medical image analysis
community of using RFs for handling complex 
classification/regression tasks including
normal brain segmentation \citep{yi2009},
MS lesion segmentation \cite{geremia2011}, 
multimodal brain tumor segmentation
\citep{bauer2012,zikic2012}, brain extraction \citep{iglesias2010}, 
anatomy detection in computed tomography \citep{criminisi2013}, and
segmentation of echocardiographic images \citep{verhoek2011}. 
A thorough introduction for those interested in delving deeper 
into the more theoretical aspects of RFs can be found
in \cite{criminisi2011}.

@@

Instrumental to our success in the 2013 challenge 
was the use of image features based on multimodal shape and appearance 
asymmetries which were generated via the construction of symmetric, 
multivariate brain templates.  Additional improvements in performance
were due to the use of a two-stage RF model approach
whereby the output probability images from the application of the first
RF model seeds the construction of a subset of new feature images
for input into the second stage.  The results of the second RF model
are then refined using a series of heuristically-derived binary morphological
operations.  The two-stage approach is motivated by
the fact that the RF is a powerful tool for ranking the features
in a high-dimensional model i.e. can be used for model selection.
However, the RF outputs are not geometrically constrained, which is
the motivation for our second refinement step.

In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
we have made the code available as open source%
\footnote{
https://github.com/ntustison/BRATS2013
}
and, to our knowledge, we are the only competitors out of the approximately 20 teams 
that participated in either years' challenges to do this.  All code is based
on the well-vetted and open source Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
} 
of the National Institutes of Health and is available through our \textit{ANTsR} 
package.
In subsequent sections we describe in greater detail the automated 
supervised brain segmentation pipeline including the generation of asymmetry
feature images and the concatenated RF construction.  We also
describe the code and data used for the BRATS challenge so that 
the interested reader can reproduce our results.  Finally, we 
illustrate performance with results from the BRATS 2013 challenge.

@@

Certain key elements characterize the proposed supervised brain tumor segmentation
protocol including:
\begin{itemize}
  \item construction of symmetric multivariate templates for multimodal 
        asymmetry-based features,
  \item generation of other image-based features,
  \item training and geometric refinement of RF models, and
  \item prediction using the proposed framework.
\end{itemize}
In subsequent sections, we describe these items and detail the
\textit{ANTsR} framework which coordinates all steps.

\subsection{Image Features for RF Model Training and Prediction}
Feature generation is a critical step in machine learning regardless
of which prediction methods are employed.  Below, we detail the
feature extraction approach and how these results are fed into the
random forest technology on which we based our final segmentations.

\subsubsection{Multimodal Asymmetry Features from Multivariate Symmetric Templates}

@@

\begin{figure*}
    \centerline{\includegraphics[width=130mm]{templateKirby.pdf}}
  \caption{Multivariate symmetric template created from the MMRR 
           data \citep{landman2011}.  Of the seven modalities 
           comprising the set of study acquisitions, we illustrate the
           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
           Although DWI-based images were not included in the challenge data, 
           such images have shown discriminative potential \citep{price2003,cha2005} 
           warranting investigation in our future work.  
           The optimal transformation and averaging of the individual 
           subject images result in the asymmetric template represented at 
           the top of the middle column.  A horizontal reflection 
           perpendicular to the mid-sagittal
           plane, $T_R$, resulted in the contralateral counterpart represented
           at the bottom of the second column.  The final template seen on
           the right is a result of repeating the template construction using
           the two asymmetric templates as input.
          }
  \label{fig:symmetrictemplates}
\end{figure*}

To take into account these potential asymmetries, we 
require a data set with the same modalities as dictated
by the subject image acquisition protocol.  Although it
would be preferable to build population-specific multivariate
templates from normal data using the same acquisition 
parameters \citep{avants2010}, we substituted well-known,
publicly available data since such normal data was not made available
for the BRATS 2013 challenge. 
A recent neuroimaging reproducibility study
by Landman et al. resulted in an open data cohort of 21
normal individuals, each imaged twice, comprising several
modalities including ASL, FLAIR, DTI, fMRI, T1, and T2 
\citep{landman2011}.  These data (known as the
``MMRR'' data set) were selected for deriving
a multivariate template due to its public availability and
inclusion of several modalities (even permitting future 
incorporation of modalities 
not currently included with the BRATS challenge into our 
segmentation framework).

\begin{figure}
  \centering
%    \includegraphics[width=80mm]{asymmetryFeatureImages.pdf}
    \includegraphics[width=120mm]{asymmetryFeatureImages.pdf}
  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.  Feature images used are outlined in a dashed line.
Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
          }
  \label{fig:asymmetryFeatures}
\end{figure}

As detailed in \cite{avants2008,avants2010}, 
given $K$ multimodality images, ${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, 
for $N$ subjects,  multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ (with corresponding 
coordinate system $\psi(\mathbf{x})$) to minimize the
following cost function:
\begin{align}
  \label{test}
  \sum_{n=1}^N \Bigg[ &D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \nonumber  \\
           &+ \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n,J_k   
           \left(\phi^{-1}_n(\mathbf{x},1)\right)\right)\Bigg]
\end{align}
%\begin{align}
%  \sum_{n=1}^N 
%        \left[ D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right)   
%        + \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\right]
%\end{align}
$D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.  Once the multivariate
template has converged (typically in four iterations), we symmetrize
the template by flipping each asymmetric template component contralaterally 
and then running the
multivariate template construction a second time using only the multivariate
template and its symmetric analog.  This is illustrated conceptually in
Figure \ref{fig:symmetrictemplates}.

@@

After constructing the template offline, each data set is processed
by first registering the non-contrast T1-weighted image to the T1-weighted
component of the symmetric template.  To do this we use a recently developed 
SyN \citep{avants2011a} variant based on B-spline regularization 
which has demonstrated good performance in normal brain registration \citep{tustison2013a}.
Good performance also extends to this pathological data scenario as 
indicated by visual inspection and the fact that the derived features
were amongst the most informative in our winning entry.

We denote the mapping from the subject, $S$, to the template, $T$,
space as $\phi_b: S \leftrightarrow \underset{b}{\leftrightsquigarrow}  T$
which consists of both affine and diffeomorphic components.  Transform
invertibility is essential for the template-based features.  The first 
set of feature images is generated by warping the template components 
to the subject space and calculating the difference image (see the 
top portion of Figure \ref{fig:asymmetryFeatures}).  For example,
the T2 symmetric template voxelwise difference image is calculated from
\begin{align}
  \mathrm{T2}\,\,\mathrm{symmetric}\,\,\mathrm{template}\,\,\mathrm{difference} = S_{\mathrm{T2}} - T_{\mathrm{T2}}\left(\phi_b^{-1}\right).
\end{align}
Note that the T1-weighted contrast and non-contrast images are paired with
the T1-weighted component of the symmetric template.  The second
set of template-based feature images is generated per modality 
as the difference image with the contralateral reflection.  This is achieved by calculating the 
reflection transform, $T_R$, in the symmetric template space and composing transforms
as follows to create the non-Euclidean contralateral counterpart per modality:
\begin{align}
  S_{contralateral} = S\left( \phi_b^{-1}\left( T_R\left( \phi_b \right)\right)\right).
\end{align}
This process is illustrated in the bottom portion of Figure \ref{fig:asymmetryFeatures}. A final, related feature is the Jacobian
of the transform.  The basis for this final feature is that 
relatively larger Jacobian values might be indicative of tumor
expansion.

\begin{figure*}
\centering
  \makebox[0.9\textwidth][c]{
  \includegraphics[width=150mm]{featureImages.pdf}
  }
  \caption{Representative feature images derived from the 
           BRATS\_HG0301  ``challenge'' data set.
           Neighborhood statistical images for each modality were 
           generated by calculating a given statistic within
           a specified neighborhood radius.  Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific 
           difference images with the
           corresponding symmetric template itself and with respect to the 
           contralateral side.   
           This mapping is also used to produce the log Jacobian image.  Finally, the 
           (T1 - TC) image is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
  \label{fig:featureImages}         
\end{figure*}

\subsubsection{Image Features for Concatenated Random Forests}

>>>>

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{K.~L.~Shrinidhi}
\author[label1]{Max Wintermark}
\author[label1]{Christopher R.~Durst}
\author[label2]{Benjamin M.~Kandel}
\author[label2]{James C.~Gee}
\author[label3]{Murray C.~Grossman}
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, 
                 Department of Radiology, University of Pennsylvania,
                Philadelphia, PA}
\address[label3]{Department of Neurology, University of Pennsylvania,
                Philadelphia, PA}

@@

\begin{abstract} 
Automated brain tumor segmentation in MRI is complicated by the lack
of prior knowledge concerning tumor location, spatial extent, shape,
possible displacement of normal tissue, and intensity signature.
%These factors limit the applicability of techniques with established
%performance in related tasks such as normal brain tissue
%segmentation.  
To accommodate such complications, we introduce a framework for supervised
segmentation based on multiple modality intensity, geometry, and asymmetry feature sets.  These features drive a supervised whole-brain and tumor segmentation
approach based on random forest-derived probabilities.  The
asymmetry-related features (based on optimal symmetric multimodal 
templates) demonstrate excellent discriminative 
properties within this framework. We also gain performance by  
generating probability maps from random forest models and using these maps 
for a refining Markov random field regularized probabilistic segmentation.  
This strategy allows us to interface the supervised learning capabilities 
of the random forest model with regularized probabilistic segmentation
using the recently developed \textit{ANTsR} 
package---a comprehensive statistical and visualization interface between the
popular Advanced Normalization Tools (ANTs) and the \textit{R}
statistical project. 
%\textit{ANTsR} integrates proven image analysis methods
%for image registration, segmentation, bias correction, and
%template building with \textit{R}'s statistical and machine
%learning techniques and state-of-the-art visualization possibilities.  
The reported algorithmic framework was the
top-performing entry in the MICCAI 2013 Multimodal Brain Tumor
Segmentation challenge.  The challenge data were widely varying 
consisting of both high-grade and low-grade glioma tumor 
four-modality MRI from five different institutions.
Average Dice overlap measures for the final algorithmic assessment
were 0.87, 0.78, and 0.74 for ``complete'', ``core'', and ``enhanced'' 
tumor components, respectively.
%and, to our knowledge, is the only entry for
%both challenge years (2012 \& 2013) which has been made publicly
%available.
\end{abstract}

@@

\section{Introduction}
Given the complexity of tumor growth and appearance and the need
for precise volumetric measurements for tumor characterization, 
much research has been invested in computational methods for 
automated segmentation of tumor regions in MR images.  Several 
approaches have been previously proposed in the literature as
detailed in recent reviews \citep{angelini2007,bauer2013}.  
Commensurable algorithmic evaluation, however, is
extremely problematic due to widely varying performance assessments
described in the corresponding publications, lack of publicly available 
evaluation data, and private algorithmic instantiations.  
In response to these issues, the Multimodal Brain Tumor Segmentation 
(BRATS) challenge was initiated in 2012 (and continued in 2013) under 
the auspices of the Medical Image Computing and Computer Assisted 
Intervention Society in association with their annual international 
conference \citep{menze2014}.

After observing the 2012 challenge to gain insight for our own research 
needs \citep{durst2014}, we found few implementation resources available.
  This prompted our participation in the challenge which took place
the following year in Nagoya, Japan concerning brain tumor segmentation
of a varied data set from five different institutions involving FLAIR, 
T1-weighted, T1-weighted contrast (T1C), and T2-weighted MRI modalities.  Heavily influenced by the success 
of participants in the 2012 challenge (specifically, the work of 
\cite{bauer2012,geremia2012,zikic2012}), we adopted the increasingly popular 
random forest (RF) machine learning framework \citep{breiman2001} which 
permits the inclusion of many potentially discriminative image features.  
Supervised segmentation techniques generally consist of a training phase
for model construction using image-based feature data followed by prediction using the 
generated model.  For supervised brain tumor segmentation, 
a set of training data consisting of labeled brain images 
is used to construct a predictive model.  Although other 
classification techniques have been used to segment
brain tumors (e.g., support vector machines \citep{bauer2011}),
RF models have proven particularly successful.

Several machine learning concepts were integrated to create 
the RF framework first articulated in its entirety by Breiman
et al. \citep{breiman2001} for performing classification/regression.  
Although decision trees had been previously explored in the literature, 
it was the success of ``boosting''-style machine learning 
techniques, such as AdaBoost \citep{schapire1990,freund1997}, which influenced 
the aggregation of such decision trees into ``forests'' 
with randomized node optimization for improved
classification/regression performance \citep{ho1995,amit1997}.
The final element of bootstrap aggregating or ``bagging'' (i.e.
random sampling of the training data) was
introduced by Breiman \citep{breiman1996} to achieve improved
accuracy.%
%\footnote{
%One of the principal advantages of \textit{R} is the extensive community of
%developers  who have contributed on the order of thousands of packages 
%extending \textit{R} 's capabilities beyond its core functionality.
%Most relevant here
%is the {\tt randomForest} package developed from Breiman's original
%Fortran code by Liaw and Wiener \citep{liaw2002}.
%}

Early adoption \citep{viola2005} and success in the
computer vision community
has led to a recent surge within the medical image analysis
community of using RFs for handling complex 
classification/regression tasks including
normal brain segmentation \citep{yi2009},
MS lesion segmentation \cite{geremia2011}, 
multimodal brain tumor segmentation
\citep{bauer2012,zikic2012}, brain extraction \citep{iglesias2010}, 
classification of Alzheimer's disease \citep{gray2013},
anatomy detection in computed tomography \citep{criminisi2013}, and
segmentation of echocardiographic images \citep{verhoek2011}. 
A thorough introduction for those interested in delving deeper 
into the more theoretical aspects of RFs can be found
in \cite{criminisi2011}.

@@

Instrumental to our success in the 2013 challenge 
was the use of image features based on multimodal shape and appearance 
asymmetries which were generated via the construction of symmetric
multivariate brain templates.  Additional improvements in performance
were due to the use of a two-stage RF model approach
whereby the output probability images from the application of the first
RF model seeds the construction of a subset of new feature images
for input into the second stage.  
%The results of the second RF model application
%are then refined using a series of heuristically-derived binary morphological
%operations.  
The two-stage approach is motivated by
the fact that the RF is a powerful tool for ranking the features
in a high-dimensional model i.e., it can be used for model selection
in voxel-wise classification.  However, the RF outputs are not geometrically 
constrained, which is the motivation for our second refinement step.

In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
we have made the code available as open source%
%\footnote{
%https://github.com/ntustison/BRATS2013
%}
and, to our knowledge, we are the only competitors out of the approximately 20 teams 
that participated in either years' challenges to do this.  All code is based
on the well-vetted and open source Insight Toolkit (ITK)%
%\footnote{
%http://www.itk.org/
%} 
of the National Institutes of Health and is available through our \textit{ANTsR} 
package.
In subsequent sections we describe in greater detail the automated 
supervised brain segmentation pipeline including the generation of asymmetry
feature images and the concatenated RF construction.  We also
describe the code and data used for the BRATS challenge so that 
the interested reader can reproduce our results.  Finally, we 
illustrate performance with results from the BRATS 2013 challenge.

@@

Feature generation for the proposed framework is critical.  The performance
of the selected machine learning strategy is limited by the input used for 
training. The image features described below were selected based on previous
work in the literature and our own observations regarding tumor characteristics
and what could possibly be encoded algorithmically.  Observations 
regarding tumor intensity differences from normal tissue has been encoded
in previously proposed research by calculating normalized neighborhood 
intensity statistics and intensity modeling \citep[e.g.,][]{bauer2011,bauer2012,geremia2012,zikic2012}.
In these approaches both the features and classification algorithm are limited to 
voxelwise considerations.

Our success in the MICCAI 2013 challenge was largely due to the inclusion of
an extremely important set of features based on the common observation of
bilateral asymmetrization caused by tumor presence.  Additionally, several
features were designed to encode spatial coherence even though the RF classification
is performed on a voxelwise basis. For example, connected component geometry of the
intensity modeling labels were used to discriminate observations of compact,
isotropic objects (vs. highly anisotropic) typical of tumor presence and growth 
\citep{greenspan1972}.  Table \ref{table:features} provides a listing of image
features used and the motivation for their inclusion.

Certain key elements characterize the proposed supervised brain tumor segmentation
protocol including:
\begin{itemize}
  \item construction of symmetric multivariate templates,
  \item generation of image features including those based on:
    \begin{itemize}
      \item asymmetry and template normalization, 
      \item intensity modeling and connected component geometry,      
      \item neighborhood first-order statistics, and
      \item brain mask coordinate systems.
    \end{itemize}
  \item training and geometric refinement of RF models, and
  \item prediction using the proposed framework.
\end{itemize}
In subsequent sections, we describe these items and detail the
\textit{ANTsR} framework which coordinates all steps.

\begin{table}
\caption{Image feature list for the proposed supervised brain tumor segmentation.}
\label{table:features}
\centering
\begin{tabular*}{0.99\textwidth}{@{\extracolsep{\fill}} c c l}
\toprule
\multicolumn{3}{c}{\bf Asymmetry and template normalization} \\
\midrule
\multicolumn{1}{c}{Feature} & \multicolumn{1}{c}{Number} & \multicolumn{1}{c}{Motivation} \\
\midrule
contralateral difference & 1 per modality & contralateral comparison \\
template difference  & 1 per modality & average normal comprison\\
log Jacobian &  1 & tumor distortion \\
warped template distance & 1 & tumor distortion \\
\midrule
\multicolumn{3}{c}{\bf Intensity modeling and connected component geometry} \\
\midrule
\multicolumn{1}{c}{Feature} & \multicolumn{1}{c}{Number} & \multicolumn{1}{c}{Motivation} \\
\midrule
Pr(CSF)  & 1 per modality & normal tissue intensity\\
Pr(gray matter) & 1 per modality & normal tissue intensity\\
Pr(white matter) & 1 per modality & normal tissue intensity\\
Pr(edema) & 1 per modality & normal tissue intensity\\
Pr(non-enhancing) & 1 per modality & tumor tissue intensity\\
Pr(enhancing) & 1 per modality & tumor tissue intensity\\
Pr(necrosis) & 1 per modality & tumor tissue intensity\\
elongation & 1 per modality & anisotropic components \\
eccentricity & 1 per modality & anisotropic components \\
volume & 1 per modality & small, isolated components \\
distance to tumor core & 1 per modality & proximity to tumor core \\
$\frac{\mathrm{volume}}{\mathrm{surface\,\,area}}$ & 1 per modality & anisotropic components \\
\midrule
\multicolumn{3}{c}{\bf Neighborhood first-order statistics} \\
\midrule
\multicolumn{1}{c}{Feature} & \multicolumn{1}{c}{Number} & \multicolumn{1}{c}{Motivation} \\
\midrule
mean (radius = 1)  & 1 per modality &  tissue \& tumor intensity \\
std. dev. (radius = 1)  & 1 per modality & tissue \& tumor  intensity \\
skewness (radius = 1)  & 1 per modality & tissue \& tumor  intensity \\
mean (radius = 3)  & 1 per modality & tissue \& tumor  intensity \\
std. dev. (radius = 3)  & 1 per modality & tissue \& tumor  intensity \\
skewness (radius = 3)  & 1 per modality & tissue \& tumor  intensity \\
T1, T1C difference & 1 & isolate tumor tissue \\
\midrule
\multicolumn{3}{c}{\bf Brain mask coordinate system} \\
\midrule
\multicolumn{1}{c}{Feature} & \multicolumn{1}{c}{Number} & \multicolumn{1}{c}{Motivation} \\
\midrule
subject distance & 1 & peripheral vs. subcortical \\
\bottomrule
\end{tabular*}
\end{table}

%FeatureSet ,  Number of Features ,   Motivation  ,  Modalities involved
%
%Asymmetry ,       2                        ,    ....            ,

\subsection{Optimal Symmetric Multivariate Templates}

@@

\begin{figure*}
    \centerline{\includegraphics[width=130mm]{templateKirby.pdf}}
  \caption{Multivariate symmetric template created from the MMRR 
           data \citep{landman2011}.  Of the seven modalities 
           comprising the set of study acquisitions, we illustrate the
           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
           Although DWI-based images were not included in the challenge data, 
           such images have shown discriminative potential \citep{price2003,cha2005} 
           warranting investigation in our future work.  
           The optimal transformation and averaging of the individual 
           subject images result in the asymmetric template represented at 
           the top of the middle column.  A horizontal reflection, $T_R$,
           perpendicular to the mid-sagittal plane  resulted in the 
           contralateral counterpart represented
           at the bottom of the second column.  The final template seen on
           the right is a result of repeating the template construction using
           the two asymmetric templates as input.
          }
  \label{fig:symmetrictemplates}
\end{figure*}

To take into account these potential asymmetries and the distortion of 
the normal plane of symmetry in the brain, we 
require a data set with the same modalities as dictated
by the subject image acquisition protocol.  Although it
is preferable to build population-specific multivariate
templates from normal data using the same acquisition 
parameters \citep{avants2010}, such data were not available
from the BRATS challenge.  Therefore, we substituted well-known,
publicly available data from a recent neuroimaging reproducibility study
by Landman et al. which resulted in an open data cohort of 21
normal individuals, each imaged twice, comprising several
modalities including ASL, FLAIR, DTI, fMRI, T1, and T2 
\citep{landman2011}.  These data (known as the
``MMRR'' data set) were selected for deriving
a multivariate template due to its public availability and
inclusion of several modalities even permitting future 
incorporation of modalities 
not currently included with the BRATS challenge into our 
segmentation framework.

\begin{figure}
  \centering
%    \includegraphics[width=80mm]{asymmetryFeatureImages.pdf}
    \includegraphics[width=120mm]{asymmetryFeatureImages.pdf}
  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.  $T_R$ denotes a horizontal reflection
  perpendicular to the mid-sagittal plane.
  Feature images used are specified by the dashed box.
Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
          }
  \label{fig:asymmetryFeatures}
\end{figure}

As detailed in \cite{avants2008,avants2010}, 
given $K$ multimodality images, ${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, 
for $N$ subjects,  multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$, with corresponding 
coordinate system $\psi(\mathbf{x})$, to minimize the
following cost function:
\begin{align}
  \label{test}
  \sum_{n=1}^N \Bigg[ &D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \nonumber  \\
           &+ \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n,J_k   
           \left(\phi^{-1}_n(\mathbf{x},1)\right)\right)\Bigg].
\end{align}
%\begin{align}
%  \sum_{n=1}^N 
%        \left[ D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right)   
%        + \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\right]
%\end{align}
$D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.  Once the multivariate
template has converged (typically in four iterations), we symmetrize
the template by flipping each asymmetric template component contralaterally 
and then running the
multivariate template construction a second time using only the multivariate
template and its symmetric analog.  This is illustrated conceptually in
Figure \ref{fig:symmetrictemplates}.

@@

\subsection{Image Features Based on Asymmetry and Symmetric Template Normalization}

After constructing the template offline, each data set is processed
by first registering the non-contrast T1-weighted image to the T1-weighted
component of the symmetric template.  To do this we use a recently developed 
SyN \citep{avants2011a} variant based on B-spline regularization 
which has demonstrated good performance in normal brain registration \citep{tustison2013a}.
%Good performance also extends to this pathological data scenario as 
%indicated by visual inspection and the fact that the derived features
%were amongst the most informative in our winning entry.
We denote the mapping from the subject, $S$, to the template, $T$,
space as $\phi_b: S \leftrightarrow \underset{b}{\leftrightsquigarrow}  T$
which consists of both affine and diffeomorphic components.  Note that transform
invertibility is essential for the template-based features.  The first 
set of feature images is generated by warping the template components 
to the subject space and calculating the difference image (see the 
top portion of Figure \ref{fig:asymmetryFeatures}).  
Note that the T1-weighted contrast and non-contrast images are paired with
the T1-weighted component of the symmetric template.
For example,
the T2 symmetric template voxelwise difference image is calculated from
\begin{align}
  \mathrm{T2}\,\,\mathrm{symmetric}\,\,\mathrm{template}\,\,\mathrm{difference} = S_{\mathrm{T2}} - T_{\mathrm{T2}}\left(\phi_b^{-1}\right).
\end{align}

The second
set of template-based feature images is generated per modality 
as the difference image with the contralateral reflection.  This is achieved by calculating the 
reflection transform, $T_R$, in the symmetric template space and composing transforms
as follows to create the non-Euclidean contralateral counterpart per modality:
\begin{align}
  S_{contralateral} = S\left( \phi_b^{-1}\left( T_R\left( \phi_b \right)\right)\right).
\end{align}
The corresponding feature image is calculated as the voxelwise difference $S_{contralateral} - S$.
This process is illustrated in the bottom portion of Figure \ref{fig:asymmetryFeatures}. Two 
related features are the Jacobian determinant image 
of the transform (calculated voxelwise) from the template to the subject, i.e., $J(\phi_b^{-1})$
and the warped signed distance template mask \citep{maurer2003}.
The motivation for inclusion of these two feature is that 
relatively larger Jacobian values are a potential indicator of tumor
expansion similar to the distorted distance map.

\begin{figure*}
\centering
  \makebox[0.9\textwidth][c]{
  \includegraphics[width=150mm]{featureImages.pdf}
  }
  \caption{Representative feature images derived from the 
           BRATS\_HG0301  final assessment data set.
           Neighborhood statistical images for each modality were 
           generated by calculating a given statistic within
           a specified neighborhood radius.  Similarly, we 
           calculate the voxelwise difference image between the
           T1 and T1C image.  
           of 
           Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific 
           difference images with the
           corresponding symmetric template itself and with respect to the 
           contralateral side.   This mapping is also used to produce the log Jacobian image
           and warped template distance mask.  Additionally, the 
           (T1 - T1C) image is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
  \label{fig:featureImages}         
\end{figure*}
